---
title: "New York City's Taxi and Limousine Commission (TLC) Trip Data"
author: "Wencong (Priscilla) Li & Trang Le"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NYCTAXI DATA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## NYC Taxi

NYC's [Taxi and Limousine Commission Trip Data](http://www.nyc.gov/html/tlc/html/about/trip_record_data.shtml) is a collection of trip records including fields capturing pick-up and drop-off locations, times, trip distances, fares, rate types, and driver-reported passenger counts. The data was collected and provided to the NYC TLC by technology providers under the Taxicab & [Livery Passenger Enhancement Programs](http://www.nyc.gov/html/tlc/html/industry/shl.shtml). 

## Getting started
To start using the package, load the 'nyctaxi' package into your R session. Since the 'nyctaxi' package currently lives on GitHub and not on CRAN, you have to install it using 'devtools'.

```{r}
#install.packages("devtools")
#devtools::install_github("beanumber/nyctaxi")
```

## Using the NYC Taxi Trip Data
Two dataframes are included in this package: 'green_2016_01_sample' and 'yellow_2016_01_sample'. There are random samples of 1000 observations generated by the 'sample' function in base R from the 2016 January green and yellow taxi trip data.

```{r, warning=FALSE, message=FALSE}
library(nyctaxi)
data(green_2016_01_sample)
head(green_2016_01_sample)
```

## Extracting, Transforming and Loading the data

To access data during wider time spans, make use of the ['etl' package](https://github.com/beanumber/etl) to download the data and import it into a database. Please see the documentation for 'etl_extract' for further details and examples.

```{r, warning=FALSE, message=FALSE}
help("etl_extract.etl_nyctaxi")
```

The code below creates a directory on your local desktop and downloads NYC taxicab trip data from Janaury, 2016 to your local directory. It also transforms/cleans the data and loads it to a sqlite database.  
```{r, eval=FALSE, warning=FALSE, message=FALSE}
taxi <- etl("nyctaxi", dir = "~/Desktop/nyctaxi/")

taxi %>%
  etl_extract(years = 2016, months = 1, types = c("green")) %>% 
  etl_transform(years = 2016, months = 1, types = c("green")) %>% 
  etl_load(years = 2016, months = 1, types = "green")}
```

## Using the NYC Green Taxi Trip Data

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(leaflet)
library(lubridate)
```

We can use `leaflet` to visualize the pickup and dropoff locations of the 1000 trips in the green taxi trip dataset:
```{r, warning=FALSE, message=FALSE}
my_trips <- green_2016_01_sample

#clean_up data according to date and time of pickup
one_cab <- my_trips %>% 
  filter(Pickup_longitude != 0)

leaflet(data = one_cab) %>% 
  addTiles() %>% 
  addCircles(lng = ~Pickup_longitude, lat = ~Pickup_latitude) %>% 
  addCircles(lng = ~Dropoff_longitude, lat = ~Dropoff_latitude, color = "green")
```


We can use `lubridate` to clean datetime variable:
```{r, warning=FALSE, message=FALSE}

clean_datetime <- my_trips %>% 
  mutate(lpep_pickup_datetime = ymd_hms(lpep_pickup_datetime)) %>%
  mutate(Lpep_dropoff_datetime = ymd_hms(Lpep_dropoff_datetime)) %>% 
  mutate(weekday_pickup = weekdays(lpep_pickup_datetime)) %>%
  mutate(weekday_dropoff= weekdays(Lpep_dropoff_datetime))
```

We can now analyze the number of trips occurred on each day of a week:
```{r, warning=FALSE, message=FALSE}
clean_datetime %>% 
  group_by(weekday_pickup) %>%
  summarize(N = n(), avg_dist = mean(Trip_distance), 
            avg_passengers = mean(Passenger_count), 
            avg_price = mean(Total_amount))
```
It looks like on Friday and Saturday had the most trips.

